---
title: "Corporate Financial Management"
subtitle: "Day 2 Morning: Sources of Finance"
author: "Prof. Barry Quinn"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(ggplot2)
```

## Sources of Finance Framework {.unnumbered}

This morning, we build upon the capital investment decisions covered yesterday afternoon to focus on the financing decision - how to source appropriate capital to fund investment opportunities identified through investment appraisal techniques.

:::{.callout-note}
## Learning Outcomes for This Session
By the end of this morning, you will be able to:

1. Understand the financing decision context and its relationship to investment decisions
2. Identify and evaluate the major internal and external sources of finance
3. Assess the benefits and drawbacks of different financing methods
4. Explain the different approaches to obtaining finance through intermediation and disintermediation
5. Apply knowledge to determine appropriate financing strategies for organisations of different sizes
:::

## The Financing Decision

After making capital investment decisions using the appraisal techniques covered yesterday, the finance manager must turn attention to the financing decision – finding suitable sources to fund those investments.

:::{.panel-tabset}
### The Financing Context

The financing decision follows logically from the investment decision and involves determining:

- What sources of finance are available (internal vs external)
- What term of borrowing the funding is available for
- How much each potential source of finance costs (cost of capital)

### Financing Decision Framework

```{mermaid}
flowchart TD
    A[Investment Decision Made] --> B[Identify Financing Needs]
    B --> C[Review Internal Sources]
    B --> D[Review External Sources]
    C --> E[Evaluate Cost and Availability]
    D --> E
    E --> F[Structure Financing Package]
    F --> G[Implement Financing Strategy]
    G --> H[Monitor and Review]
```

### Real-World Context

Financing decisions are critical to:

- Enterprise sustainability and growth
- Risk management
- Shareholder value creation
- Corporate governance
- Financial flexibility
:::

## Internal vs External Sources of Finance

Organisations can source funding either internally from operations or externally from capital markets and financial institutions.

:::{.grid}
::: {.g-col-6}
**Internal Finance Sources:**

Internal finance comes from the operational activities of the business itself and includes:

- Retained earnings (profit ploughback)
- Working capital management:
  - Tighter credit control
  - Reduced inventory levels
  - Delayed payment to suppliers
:::

::: {.g-col-6}
**External Finance Sources:**

External finance is raised from outside the business through:

- Debt (borrowings)
- Equity (shares)
- Hybrid instruments (convertibles, preference shares)
- Leasing and similar arrangements
- Factoring and invoice discounting
:::
:::

## The Major Internal Sources of Finance

Internal financing is often preferred as it avoids the costs, formalities, and external scrutiny associated with raising external finance.

```{mermaid}
flowchart LR
    A[Internal Finance Sources] --> B[Short-term Sources]
    A --> C[Long-term Sources]
    B --> D[Tighter credit control]
    B --> E[Reduced inventory levels]
    B --> F[Delayed payments to creditors]
    C --> G[Retained profits]
```

:::{.callout-important}
## Working Capital Management
Effective working capital management can release significant funds from day-to-day operations. However, there are limits to such strategies:

- Excessive credit control may alienate customers
- Minimal inventory may lead to stockouts and lost sales
- Delayed supplier payments can damage critical relationships
:::

## The Major External Sources of Finance

External sources represent a wide spectrum of options ranging from short-term facilities to long-term capital.

```{mermaid}
flowchart TD
    A[External Finance Sources] --> B[Short-term: < 1 year]
    A --> C[Medium-term: 1-7 years]
    A --> D[Long-term: > 7 years]
    B --> E[Bank overdrafts]
    B --> F[Invoice discounting]
    B --> G[Debt factoring]
    B --> H[Bills of exchange]
    C --> I[Term loans]
    C --> J[Hire purchase]
    C --> K[Leasing]
    D --> L[Corporate bonds]
    D --> M[Preference shares]
    D --> N[Ordinary shares]
    D --> O[Securitisation]
```

## Approaches to Obtaining Finance

There are two primary approaches through which organisations can access external finance.

:::{.panel-tabset}
### Intermediation

**Intermediation** is where finance is arranged through financial institutions such as commercial banks.

**Characteristics:**
- Institution acts as intermediary between savers and borrowers
- Involves screening and monitoring by the institution
- Offers professional expertise and risk assessment
- Typically used by businesses of all sizes
- Often involves developing ongoing banking relationships

**Examples:**
- Bank loans and overdrafts
- Factoring and invoice discounting
- Leasing arrangements through finance companies

### Disintermediation

**Disintermediation** is where the financial intermediary's role is bypassed, with companies going directly to the money markets or stock exchanges to obtain finance.

**Characteristics:**
- Direct access to capital markets
- Restricted to large, reputable companies
- Used for substantial funding requirements
- Often more cost-effective for large sums
- Requires substantial disclosure and market reputation

**Examples:**
- Corporate bond issues
- Public share offerings
- Commercial paper issuance
:::

## Bank Finance: The Lending Decision

Banks are a primary source of external finance for many businesses. Understanding their lending criteria is essential for successful financing applications.

:::{.panel-tabset}
### The Lending Assessment

Banks typically consider two core aspects when making lending decisions:

1. **Going concern assessment**: Determining whether the borrowing company can afford to make repayments from its ongoing operations.

2. **Liquidation assessment**: Evaluating the recoverability of capital in the event the company goes into liquidation.

### The Seven C's of Lending

A traditional framework used by banks to evaluate potential borrowers includes:

1. Character
2. Capacity
3. Capital
4. Conditions
5. Customer relations
6. Competition
7. Collateral

### The Expanded View: 20 C's of Credit

```{r}
c_categories <- data.frame(
  Category = 1:20,
  Description = c("Character or culture of the borrower",
                 "Competence of borrower",
                 "Continuity of management",
                 "Corporate constitution",
                 "Customers and competitors",
                 "Controls of costs and cash",
                 "Capacity to contract",
                 "Credibility of accounts",
                 "Cause or purpose and term of loan",
                 "Capital required in total",
                 "Capital contribution of the bank",
                 "Contribution to bank profits",
                 "Commitments",
                 "Contingencies",
                 "Comprehensive cash-flow projections",
                 "Current trading",
                 "Creditor coordination",
                 "Conditions of drawdown",
                 "Continuing covenants",
                 "Collateral of security")
)

kable(c_categories, col.names = c("No.", "Description"), align = "lr")
```

### Business Plan Requirements

A strong business plan for bank finance applications should include:

- Funding requirements (amount, term, repayment, interest structure)
- Clear purpose of the finance
- Background information on the company
- Credit worthiness documentation
- Business stake and commitment
- Available security details
- Financial projections and scenario analysis
- Historical and projected financial statements
:::

## Short-Term Sources of Finance

Short-term finance options (less than one year) are typically used to support working capital needs and temporary funding requirements.

:::{.panel-tabset}
### Overview of Short-Term Finance

Short-term funding sources include:

- Working capital management
- Retained earnings
- Bank sources (overdrafts, short-term loans)
- Market sources (commercial paper)

These are typically less expensive but carry refinancing risk and may be subject to immediate recall.

### Bank Overdrafts

```{r}
overdraft_features <- data.frame(
  Feature = c("Definition", "Interest", "Security", "Legal status", "Flexibility", "Cost structure", "Risk factors"),
  Description = c("Facility allowing withdrawal beyond deposited amount to an agreed limit",
                 "Charged only on negative balance amount",
                 "Typically unsecured for smaller amounts",
                 "Legally repayable on demand",
                 "Highly flexible form of finance",
                 "Arrangement fees, review fees plus interest",
                 "Breaching limits can damage credit rating")
)

kable(overdraft_features, col.names = c("Feature", "Description"), align = "lr")
```

### Short-Term Loans

Short-term bank loans (up to one year) offer more structure than overdrafts:

- Arrangement fees based on borrower's credit profile
- Pre-determined repayment schedule
- Typically secured against assets
- Lower interest rates compared to overdrafts
- Early repayment usually incurs penalties
- Less flexible than overdrafts but more certain

### Factoring and Invoice Discounting

```{mermaid}
flowchart TD
    A[Client business] -->|1. Goods supplied on credit| B[Customer]
    C[Factor] -->|3. Advances 80% to client| A
    C -->|2. Invoices customer| B
    B -->|4. Pays amount owing to factor| C
    C -->|5. Pays remaining 20% to client| A
```

**Factoring vs Invoice Discounting compared:**

```{r}
factoring_comparison <- data.frame(
  Feature = c("Service provided", "Customer awareness", "Administration", "Control over debts", "Relative cost", "Confidentiality", "Typical clientele"),
  Factoring = c("Full debt administration and collection", "Customers aware of factor involvement", "Factor handles all administration", "Factor controls collection process", "Higher cost service", "Not confidential", "Smaller businesses"),
  Discounting = c("Advance against invoices only", "Customers typically unaware", "Business maintains own records", "Business retains control", "Lower cost service", "Confidential arrangement", "Larger businesses")
)

kable(factoring_comparison, col.names = c("Feature", "Factoring", "Invoice Discounting"), align = "lll")
```
:::

## Medium-Term Finance

Medium-term finance options (1-7 years) bridge the gap between short-term borrowing and long-term capital.

:::{.panel-tabset}
### Medium-Term Options Overview

Key characteristics of medium-term finance:

- Can be utilized by all company sizes
- Particularly important for SMEs
- More accessible than long-term capital markets
- Includes term loans, hire purchase, and leasing
- Generally secured against specific assets

### Term Loans

Term loans represent a formal commitment to lend a specified amount for a defined period:

- Typically 1-7 years maturity
- Can feature fixed or variable interest rates
- Secured against business assets
- Formal repayment schedule
- May include financial covenants
- More structured than overdrafts but less rigid than bonds

### Hire Purchase

```{r}
hp_diagram <- data.frame(
  Stage = 1:5,
  Description = c("Company selling asset transfers physical asset to hiree (purchaser)",
                 "Finance house (hirer) pays full price to selling company",
                 "Hiree makes down-payment and agrees to periodic payments",
                 "Hiree makes regular rental installments to finance house",
                 "Ownership transfers to hiree upon final payment")
)

kable(hp_diagram, col.names = c("Stage", "Description"), align = "lr")
```

**Advantages of Hire Purchase:**
- Easy to arrange
- Provides quick access to assets
- Widely available
- Predictable cash flow planning
- Not repayable on demand
- Tax relief on interest
- Ultimate ownership

**Disadvantages of Hire Purchase:**
- Relatively expensive
- Inflexible arrangement
- Costly cancellation penalties
- Hiree assumes maintenance/insurance responsibilities
- Fixed payment obligation regardless of asset utilization

### Leasing

Leases provide access to assets without necessarily transferring ownership.

**International Accounting Standard 17 defines a lease as:**
> "An agreement whereby the lessor conveys to the lessee, in return for a payment or series of payments, the right to use an asset for an agreed period of time."

**Key differences between leasing and hire purchase:**

```{r}
lease_hp_diff <- data.frame(
  Feature = c("Ownership transfer", "Tax treatment", "Balance sheet treatment", "Maintenance responsibility"),
  Hire_Purchase = c("Automatically transfers on final payment", "Hiree claims capital allowances", "Asset recorded on hirer's balance sheet", "Hiree assumes maintenance responsibility"),
  Leasing = c("No automatic transfer at end of term", "Lessor claims capital allowances", "Treatment depends on lease classification", "Varies according to lease terms")
)

kable(lease_hp_diff, col.names = c("Feature", "Hire Purchase", "Leasing"), align = "lll")
```

### Sale and Leaseback

A sale and leaseback arrangement allows a company to:

1. Sell an existing asset (typically property)
2. Immediately lease it back from the buyer
3. Continue using the asset while releasing equity capital
4. Generate immediate cash flow improvement
5. Potentially achieve favorable accounting treatment

This approach can be structured either with:
- A subsidiary company (lower risk)
- An external company (potentially higher risk)
:::

## Long-Term Finance: Debt

Debt represents a significant source of long-term finance for many businesses.

:::{.panel-tabset}
### Debt Finance Benefits

Long-term debt offers several advantages:

- Cheaper and quicker to obtain than equity
- Interest is tax-deductible
- No dilution of ownership or control
- Can reduce overall WACC (weighted average cost of capital)
- Can be structured with tailored repayment schedules
- Lower risk for investors (but less potential return)

### Debt Finance Drawbacks

However, debt also poses challenges:

- Increases financial risk for the company
- Requires regular interest payments regardless of profitability
- Often secured against company assets
- May include restrictive covenants
- Can limit managerial decision-making flexibility
- May restrict future borrowing capacity

### Restrictive Covenants

Lenders frequently impose covenants to protect their interests:

```{r}
covenants <- data.frame(
  Covenant_Type = c("Security restrictions", "Financial ratio requirements", "Additional debt limitations", 
                   "Liquidity requirements", "Sinking fund provisions", "Future debt restrictions", 
                   "Dividend limitations", "Share redemption restrictions"),
  Description = c("Restrictions on sale or use of secured assets",
                 "Minimum interest cover ratio requirements",
                 "Gearing ratio constraints (debt/assets, debt/equity)",
                 "Working capital ratio minimums to protect liquidity",
                 "Requirements to set aside funds for repayment",
                 "Limitations on additional bond issues",
                 "Restrictions on dividend payments",
                 "Constraints on share buybacks")
)

kable(covenants, col.names = c("Covenant Type", "Description"), align = "ll")
```

### Corporate Bonds

Corporate bonds represent a formalized debt security that can be traded in financial markets:

- Tradable loans with defined nominal (par) value
- Feature specified coupon rates for interest payments
- May be issued with redeemable or irredeemable terms
- Typically have 7-30 year maturities (if redeemable)
- Can be issued publicly or through private placement
- May be secured or unsecured

**Key bond characteristics:**
- Nominal value (typically £/€100)
- Coupon rate (interest payment)
- Redemption value (usually par value)
- Redemption date (for redeemable bonds)
:::

## Long-Term Finance: Mezzanine Finance and Preference Shares

Between traditional debt and equity lies a range of hybrid financing instruments.

:::{.panel-tabset}
### Mezzanine Finance Overview

Mezzanine finance represents higher-risk, higher-return capital that fills the gap between senior debt and equity:

- Long-term, subordinated financing
- Higher interest rates reflecting increased risk
- Often includes equity-like features (warrants, conversion rights)
- Typically unsecured
- Used as a last resort before equity dilution
- Main types include junk bonds, convertible bonds, and preference shares

### Junk Bonds

Junk bonds (high-yield bonds) are debt securities with lower credit ratings:

- Used by companies with weaker financial positions
- Carry higher interest rates to compensate for increased risk
- Attract risk-taking investors seeking higher returns
- Subject to higher default probability
- May include more stringent covenants

### Convertible Bonds

Convertible bonds combine debt features with an option to convert to equity:

- Fixed return securities with conversion option
- Allow holders to convert to shares at predetermined dates
- Often issued when share prices appear undervalued
- Represent a deferred equity sale
- Generally offer lower interest rates than straight debt
- Reduce immediate dilution while providing potential equity upside
- Potentially eliminate cash repayment requirement if converted

### Preference Shares

Preference shares combine features of both debt and equity:

**Key characteristics:**
- Part of share capital but typically not considered pure equity
- May be issued at premium above nominal value
- Usually carry fixed dividends
- Generally confer no voting rights
- May be redeemable and/or convertible
- Higher in capital structure than ordinary shares but below bonds
- Dividends are not tax-deductible

**Types of preference shares:**

```{r}
pref_share_types <- data.frame(
  Type = c("Irredeemable preference shares", "Redeemable preference shares", "Cumulative preference shares", 
          "Non-cumulative preference shares", "Participating preference shares", 
          "Convertible preference shares"),
  Description = c("Fixed dividend with no intended redemption or conversion",
                 "Issued for a predetermined period of time",
                 "Unpaid dividends accumulate as an obligation",
                 "Unpaid dividends are lost if not declared",
                 "Holders can participate in surplus profits",
                 "Convertible to ordinary shares under specified conditions")
)

kable(pref_share_types, col.names = c("Type", "Description"), align = "ll")
```

**Advantages of preference shares:**
- Dividends can be waived in difficult periods
- Preserve ownership structure
- Can improve gearing after conversion (if convertible)
- Easier to obtain than bonds for highly geared companies

**Disadvantages of preference shares:**
- More expensive than debt
- Dividends not tax-deductible
- May require redemption or conversion in future
:::

## Long-Term Finance: Equity

Ordinary shares (common stock) represent the fundamental risk capital of the business.

:::{.panel-tabset}
### Equity Fundamentals

Equity represents ownership interest in a business:

- Shareholders own a portion of the company
- Nominal value assigned but not related to market value
- May be traded (public companies) or not (private companies)
- Usually entitles holders to dividends from profits
- Provides pre-emptive rights in future share issues
- Most risky form of financing from investor perspective
- Least risky from company perspective

### Equity Advantages

From the company's perspective, equity offers several benefits:

- Lowest financial risk of all external finance
- No obligation to pay dividends
- No requirement to redeem capital
- No fixed repayment schedule
- Provides permanent capital base
- Increases capacity to raise additional debt

### Equity Disadvantages

However, equity also has significant drawbacks:

- Expensive to issue (5-10% of funds raised)
- Minimum fixed costs limit accessibility to larger companies
- Substantial administrative and regulatory requirements
- Dividends not tax-deductible
- Dilution of control
- Highest required return of all financing sources

### Methods of Issuing Equity

```{r}
equity_methods <- data.frame(
  Method = c("Initial Public Offering (IPO)", "Rights Issues", "Private Placement"),
  Description = c("First sale of shares to the public with stock exchange listing",
                 "Offering of new shares to existing shareholders in proportion to holdings",
                 "Sale of shares directly to specific institutional investors")
)

kable(equity_methods, col.names = c("Method", "Description"), align = "ll")
```

### Rights Issues

For existing listed companies, rights issues provide an efficient mechanism to raise additional equity:

- Offered to existing shareholders in proportion to holdings
- Typically priced at 15-20% discount to market value
- Lower transaction costs than public offerings
- Administratively simpler with existing shareholder base
- Existing shareholders can either subscribe or sell rights
- Market price adjusts to account for discount
:::

## The Finance Gap and Venture Capital

Many growing businesses face a "finance gap" where conventional financing sources are insufficient.

:::{.panel-tabset}
### The Finance Gap

The finance gap (or equity gap) typically emerges when:

- New companies require substantial scale of financing
- Existing small companies experience rapid growth
- Bank lending reaches prudential limits
- Retained earnings are insufficient
- Private equity holders lack capacity for further investment
- Company is too small for public market access

### Venture Capital

Venture capital provides specialized equity financing to bridge this gap:

- Focuses on high-risk, high-return, high-growth potential companies
- Typically targets 30%+ annual returns
- Portfolio approach with mixed outcomes expected
- Generally involves deals exceeding £250,000
- Requires equity stake and element of control
- Prefers companies with established track records
- Necessitates clear exit strategy (IPO or trade sale)

**Benefits to recipient companies:**
- Access to capital at crucial growth stage
- Business expertise and mentoring
- Industry contacts and networks
- Strategic guidance
- Potential follow-on financing

**Drawbacks for recipient companies:**
- Dilution of ownership
- Loss of some decision-making autonomy
- Increased reporting and accountability
- Pressure to achieve growth targets
- Exit timeline constraints

### Government Initiatives

Various government schemes aim to address the finance gap:

**Enterprise Investment Scheme (EIS):**
- Tax relief (30%) for investors in qualifying companies
- Investment limit of £1 million per annum
- Companies can raise up to £5 million annually
- Aimed at unlisted companies

**Venture Capital Trusts (VCTs):**
- Similar tax benefits to EIS
- Minimum five-year investment period
- Professionally managed funds
- Diversified portfolio approach
:::

## New Sources of Finance

The financing landscape continues to evolve with innovative funding mechanisms.

:::{.panel-tabset}
### Crowdfunding

Crowdfunding has emerged as a significant alternative funding source:

- Peer-to-peer lending platforms
- Equity-based crowdfunding
- Reward-based crowdfunding
- Donation-based funding
- Real estate crowdfunding

These platforms connect entrepreneurs directly with numerous small investors, bypassing traditional financial intermediaries.

### Asset-Based Finance

Asset-based lending focuses on the value of specific assets:

- Supply chain finance
- Marketplace lending
- Invoice trading platforms
- Peer-to-peer secured lending
- Asset refinancing

### Fintech Innovations

Financial technology is transforming the financing landscape:

- Blockchain-based financing
- Initial Coin Offerings (ICOs)
- Security Token Offerings (STOs)
- Artificial intelligence in lending decisions
- Open banking initiatives
:::

## Key Source of Finance Principles

Several fundamental principles should guide financing decisions in practice.

:::{.callout-note appearance="minimal"}
- The type of finance should match the term of the investment (matching principle)
- Management typically follows a "pecking order": internal sources first, then debt, then equity
- Market-sourced finance is primarily viable for larger companies due to transaction costs
- SMEs should prioritize bank financing, medium-term instruments, and venture capital
- Cost of each source correlates with the risk borne by investors
- Venture capital targets high-risk, high-return companies in the finance gap
- Comprehensive business plans with robust financial projections are essential for external financing
- Innovative financing sources continue to expand options, particularly for early-stage companies
:::

## For This Afternoon {.unnumbered}

This afternoon, we will build on this morning's exploration of financing sources to focus on:
- Cost of capital calculations
- Weighted Average Cost of Capital (WACC)
- Capital structure decisions
- Optimal gearing levels
- Dividend policy considerations

Please review the case studies in the module materials to prepare for our practical session.

## Additional Resources {.unnumbered}

**Core Reading:**
- Atrill, P. (2019) *Financial Management for Decision Makers*. 9th ed. London: FT Prentice Hall, Chapters 6-8
- Arnold, G. (2020) *Corporate Financial Management* (6th ed.), Chapters 8-10

**Online Resources:**
- [Bank of England Monetary Policy Reports](https://www.bankofengland.co.uk/monetary-policy-report)
- [UK Finance](https://www.ukfinance.org.uk/data-and-research)
- [British Business Bank](https://www.british-business-bank.co.uk/finance-hub/)

## Practice Questions {.unnumbered}

1. ABC Ltd is considering expanding its operations and requires £5 million in financing. Evaluate the most appropriate sources of finance for the company, assuming it is a medium-sized manufacturer with strong but fluctuating cash flows.

2. Compare and contrast factoring and invoice discounting as sources of short-term finance. Under what circumstances would you recommend each option?

3. XYZ plc is considering issuing either (a) £10 million in 10-year corporate bonds at 4.5% or (b) £10 million in preference shares with a 6% dividend. Discuss the advantages and disadvantages of each option, including their impact on the company's capital structure.

4. Discuss how the "pecking order theory" explains observed financing behaviors in practice. What factors might cause a company to deviate from this hierarchy?

5. A technology startup has exhausted its founders' personal investments and bank lending capacity but requires an additional £2 million to bring its product to market. Evaluate the potential financing options available to the company.